Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2013-12-17 paper.Rnw
cleaned up use of SNT and Gold Standard

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has  use/used find/found
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: prep
% Prepares the data in a the form needed for plotting.
<<ADNI1-xval-prep,dependson='ADNI1-xval-load',include=FALSE,cache=TRUE>>=
jens_atlases = read.csv('data/jens_atlases.csv')
jens_xval_data  <- read.csv(gzfile('data/cache/ADNI-JENS-XVAL:all_data.csv.gz'))
jens_xval_mean  <- read.csv(gzfile('data/cache/ADNI-JENS-XVAL:all_data_mean.csv.gz'))
jens_xval_mean$reg_method = factor(jens_xval_mean$reg_method, c("ANTS", "ANIMAL"))
jens_xval_mean$method.mb = factor(jens_xval_mean$method.mb, c("Majority Vote", "Cross-correlation Vote", "NMI Vote"))
mb.best.kappa = mean(subset(jens_xval_mean, reg_method   == "ANTS" & 
                                  method.mb    == "Majority Vote" &
                                  templates.mb == 19 & 
                                  atlases      == 9)$k.mb)
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% shortcuts 
\newcommand{\mb}{MAGeT-Brain}
\newcommand{\ants}{ANTS}
\newcommand{\animal}{ANIMAL}
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T}
\newcommand{\FEPdataset}{SZ-FEP}
\newcommand{\gs}{Gold Standard}
\newcommand{\fsl}{FSL}
\newcommand{\freesurfer}{FreeSurfer}
\newcommand{\maper}{MAPER}

% title and authors
%\title{Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using
%Multiple Automatically Generated Templates}
\title{Bootstrapping Multi-atlas Segmentation Using Multiple Automatically
Generated Templates for the Segmentation of the Whole Hippocampus and
Subfields}
\author[1]{Jon Pipitone}
\author[1]{Matt Min Tae Park}
\author[1]{Julie Winterburn}
\author[1,9]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4]{Jens C. Pruessner}
\author[4,5]{Martin Lepage}
\author[1,6,9]{Aristotle N. Voineskos}
\author[1,6,7,8]{M. Mallar Chakravarty}
\author[ ]{the Alzheimer's Disease Neuroimaging Initiative\footnote{Data
used in preparation of this article were obtained from the Alzheimer's Disease
Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). As such, the
investigators within the ADNI contributed to the design and implementation of
ADNI and/or provided data but did not participate in analysis or writing of this
report. A complete listing of ADNI investigators can be found at:
http://adni.loni.ucla.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health Laboratory, Hospital for Sick
Children, Toronto, ON, Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Douglas Mental Health University Institute, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, McGill University, Montreal, QC, Canada}
\affil[6]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[7]{Institute of Biomaterials and Biomedical Engineering, University of
Toronto, Toronto, ON, Canada}
\affil[8]{Rotman Research Institute, Baycrest, Toronto, ON, Canada}
\affil[9]{Institute of Medical Science, University of Toronto, Toronto, ON, Canada}

@@

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that multi-atlas approaches improve segmentation accuracy and
precision over regular atlas-based approaches.  These approaches often rely on
a large number of such manually segmented atlases (e.g.  30-80) that take
significant time and expertise to produce. We present an algorithm, \mb{} (
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar accuracy to
multi-atlas approaches. Thus, our method acts as an accurate multi-atlas
approach when using special, hard-to-define atlases that are laborious to
construct. \\
\textbf{Method:} \mb{} works by propagating atlas segmentations to a template library,
formed from a subset of target images, via transformations estimated by non-linear
image registration. The resultant segmentations are then propagated to each
target image and fused using a label fusion method. We conduct a 10-fold Monte
Carlo cross-validation of whole hippocampal segmentation on a pool of 69 ADNI
subjects over a range of parameter settings, and a leave-one-out
cross-validation (LOOCV) of hippocampal subfield segmentation using five
high-resolution atlases. Two final experiments assess \mb{} when applied to
first episode psychosis and Alzheimer's disease populations, and \mb{}
segmentations are compared with existing automated methods (\fsl{}, \freesurfer{},
\maper{}) and biases are explored.\\
\textbf{Results:} Using nine atlases and 19 template images, \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) of 0.84 over 10-folds of Monte Carlo
cross-validation on 69 subjects relative to semi-automated \gs{} segmentations,
and shows significantly lower variability in DSC than multi-atlas segmentation.
In a LOOCV, \mb{} reproduces the volumes of the cornu ammonis (CA) 1; CA4/dentate
gyrus (DG); and strata radiatum (SR), strata lacunosum (SL), and strata
moleculare (SM) hippocampal subfields with a percent error in volume that is
at or lower than that produced by image resampling. \mb{} produces hippocampal
volumes in a first episode psychosis patient population that are highly
correlated with expert manual segmentation volumes (Pearson $r = 0.877, t =
16.244, p < 0.001$). Compared to \fsl{} and \freesurfer{}, \mb{} shows much smaller
fixed volume bias (within $250 mm^3$ on average) to \gs{} segmentations available
from ADNI, as well as a conservative, rather than exaggerated, proportional
volume bias.\\ \textbf{Conclusion:} We demonstrate that \mb{} produces accurate
hippocampal segmentations using only 5 atlases over different hippocampal
definitions, disease populations, and acquisition types, as well as showing
that accurate identification of the hippocampal subfields is possible.
\end{abstract}

@@

More recently, many groups have used multiple atlases to improve overall
segmentation accuracy (i.e. multi-atlas segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into the definitive segmentation for the target. In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such as cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations. For example \citet{Collins2010} demonstrated that only
14 atlases, selected based on highest similarity between medial temporal lobe
neuroanatomy as evaluated by normalized mutual information \citep{Studholme1999}
from a library of 80 atlases, were required to achieve accurate segmentations of
the hippocampus. Additionally, several methods have been explored for label
fusion, including the STAPLE algorithm (Simultaneous Truth And Performance Level
Estimation; \citet{Warfield2004}) that computes a probabilistic segmentation
using an expectation-maximization framework from an set of competing
segmentations, or others where a subset of segmentations can be estimated using
metrics such as the sum of squared differences in the regions of interest to be
segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library ranging between 30
\citep{Heckemann2006} and 80 \citep{Collins2010} manually segmented atlases.
This strategy has the main drawback of being inflexible as it does not easily
accommodate varying the definition of the hippocampal anatomy (such as the
commonly used heuristic of subdividing the hippocampus into head, body, and tail
\citep{Poppenk2011,Pruessner2000}). Furthermore, none of these methods have
demonstrated sufficient flexibility to accommodate atlases that are somehow
exceptional such as those derived from serial histological data
\citep{Chakravarty2006,Yelnik2007} or high-resolution MRI data that enables
robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}. Due
to the recent availability of the latter, there has been increased interest in
the use of probabilistic methods for the identification of the hippocampal
subfields on standard T1-weighted images. Our group recently demonstrated that
through use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{MallarChakravarty2012}.
The novelty of this manuscript is the extension of our multi-atlas methodology
to the hippocampus using more than a single input atlas, while simultaneously
limiting the number of inputs used during segmentation, and demonstrating that
accurate identification of the hippocampal subfields is indeed possible using
this methodology.

Of special relevance to the present work is the LEAP algorithm (Learning
Embeddings for Atlas Propagation; \citet{Wolz2010}) because of its focus on
performing multi-atlas segmentation with a limited number of input atlases.  The
LEAP algorithm is a clever modification to the basic multi-atlas strategy in
which an atlas library is grown, beginning with a set of manually labelled
atlases, by successively incorporating unlabelled target images once they
themselves have been labelled using multi-atlas techniques. The sequence in
which target images are labelled is chosen so that the similarity between the
atlas images and the target images is minimised at each step, effectively
allowing for deformations between very dissimilar images to be broken up into
sequences of smaller deformations. Although \citet{Wolz2010} begin with an atlas
library of 30 MR images, this method could theoretically work using a much
smaller atlas library.  In their validation, LEAP was used to segment the whole
hippocampus in the ADNI-1 baseline dataset, achieving a mean Dice score of 0.85
against \gs{} segmentations.

@@

In this paper we have describe a thorough validation of the \mb{} algorithm for
segmentation of the hippocampus and its subfields. First, we addressed the feasibility and
accuracy of whole hippocampus segmentation with a limited number of input
atlases \citep{MallarChakravarty2012} by performing a multi-fold experiment
using a subset of the Alzheimer's Disease Neuroimaging Initiative (ADNI)
dataset over a range of parameters.  We then performed a leave-one-out
validation to determine if hippocampal subfields can be accurately identified
using our multi-atlas framework.  To ensure that we have not over-fit our
parameters to the aging or neurodegenerative brain, we also applied our
segmentations to a dataset of individuals suffering from first episode
psychosis.  Finally, validated our algorithm using all of the data available in
the \adnidataset{} sample and compare our segmentations to the other
segmentations available through the ADNI informatics portal.

@@

The simplest form of multi-atlas segmentation, which we call {\em basic 
multi-atlas segmentation}, involves three steps. First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propagated to the target image space. 
Third, the labels are combined into a single label by label fusion
\citep{Heckemann2006, Heckemann2011}. The basic multi-atlas segmentation method
is described in detail in other publications
\citep{Collins2010,Heckemann2011,Aljabar2009}. When only a single atlas is used,
basic multi-atlas segmentation degenerates into model-based segmentation: labels
are propagated from the atlas to a target, and no label fusion is needed.

\mb{} ({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates) bootstraps the creation of a large template library given a limited
input atlas library, and then uses the template library in basic multi-atlas
segmentation.  Images for the template library are selected from a set of input
target images, either arbitrarily or so as to reflect the neuroanatomy or
demographics of the target set as a whole (for instance, by sampling equally
from cases and controls).  The template library images are automatically
labelled by each of the atlases via label propagation.  Basic multi-atlas
segmentation is then conducted using the template library to segment the entire
set of target images (including the target images used in the construction of
the template library).  Since each template library image has multiple labels
(one from each atlas), the final number of labels to be fused for each target
may be quite large (i.e. \# of atlas $\times$ \# of templates).

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=\textwidth]{figure/MA-MAGeTBrain-Schematic}
  \end{centering}
  \caption{A schematic illustration of basic multi-atlas segmentation and \mb{}
  segmentation \label{fig:MAGeT}}
\end{figure}

@@

The following section describes for experiments conducted to assess the
segmentation quality of the \mb{} algorithm. The first two experiments assess the
validity of \mb{} using a cross-validation design. Experiment 1 investigates the
accuracy of \mb{} whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 assesses hippocampal subfield segmentation quality in a
leave-one-out cross-validation design. The last two experiments assess the
validity of the \mb{} algorithm when applied to different diseases: first episode
schizophrenia (Experiment 3), and Alzheimer's disease (Experiment 4).
Additionally, in Experiment 4, we compared \mb{} segmentations with those of
well-known automated methods and assessed segmentation bias.

@@

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a pool
of images and SNT \gs{} segmentations from the ADNI dataset \citep{Hsu2002}. This
form of cross-validation allows us to rigorously validate a large number of
parameter settings of \mb{} (atlas and template library sizes, registration
algorithm, and label fusion method) and select the best parameters to use in
subsequent experiments.

@@

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset{}} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Semi-automated segmentations of the
left and right whole hippocampi are made available with a subset of ADNI images
\citep{Hsu2002}. These labels have been generated using the SNT tool from
Medtronic Surgical Navigation Technologies, Louisville, CO (see Supplementary
Materials for detailed discussion of the segmentation process used). Throughout
this paper, we use these segmentations as the \gs{} for comparison.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI-Xval-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c(295,413,619,685,729,782,938,954,1018,1155,907,981,222,324,448,
                 546,553,572,602,610,814,1341,675,681,731,1130,41,68,70,101,
                 249,293,316,414,698,1206,1222,1339,751,1030,3,5,10,16,22,53,
                 168,183,241,1205,328,1095,991,213,159,337,343,642,753,1109,
                 1217,29,56,1188,1293,149,382,431,1202)

tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               jens_atlasess$RID, method = "reverse", test=FALSE, overall=TRUE)
caption = 
  "\\textbf{ADNI-1 cross-validation subset demographics.} 
  CN     - Cognitively Normal. 
  LMCI   - Late-onset Mild Cognitive Impairment. 
  AD     - Alzheimer's Disease.  
  Hisp   - Hispanic. 
  CDR-SB - Clinical Dementia Rating-Sum of Boxes.  
  ADAS   - Alzheimer's Disease Assessment Scale.  
  MMSE   - Mini-Mental State Examination."
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      label="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In this experiment, our dataset consists of 69 1.5T images and corresponding \gs{}
segmentations. In each validation round, the dataset is partitioned into a
training set consisting of images and \gs{} segmentations used as an atlas
library, and a validation set consisting of the remaining images segmented by
both \mb{} and multi-atlas. The computed segmentations were compared to the \gs{}
segmentations (see below).

@@

The automatically produces labels are compared to the \gs{} SNT labels provided by
ADNI. The segmentation accuracy reported for a parameter setting is the average
over ten validation rounds.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has  use/used find/found
%  - remove 'accuracy' unless specifically talking relative to ground truth (and
%    even then, qualify)
%  - overlap/similarity --> agreement? choose one. 
%  - check that chart labels are sensible 
%  - in particular, check that Experiment 5 match Experiment 1
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: prep
% Prepares the data in a the form needed for plotting.
<<ADNI1-xval-load,include=FALSE,cache=TRUE>>=
jens_atlases = read.csv('data/jens_atlases.csv')
jens_xval_data  <- read.csv(gzfile('data/cache/ADNI-JENS-XVAL:all_data.csv.gz'))
jens_xval_mean  <- read.csv(gzfile('data/cache/ADNI-JENS-XVAL:all_data_mean.csv.gz'))
jens_xval_mean$reg_method = factor(jens_xval_mean$reg_method, c("ANTS", "ANIMAL"))
jens_xval_mean$method.mb = factor(jens_xval_mean$method.mb, c("Majority Vote", "Cross-correlation Vote", "NMI Vote"))
mb.best.kappa = mean(subset(jens_xval_mean, 
                                  reg_method   == "ANTS" & 
                                  method.mb    == "Majority Vote" &
                                  templates.mb == 19 & 
                                  atlases      == 9)$k.mb)
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% shortcuts 
\newcommand{\mb}{MAGeT-Brain}
\newcommand{\multiatlas}{multi-atlas}
\newcommand{\ants}{ANTS}
\newcommand{\animal}{ANIMAL}
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T}
\newcommand{\FEPdataset}{SZ-FEP}
\newcommand{\fsl}{FSL}
\newcommand{\freesurfer}{FreeSurfer}
\newcommand{\maper}{MAPER}
\newcommand{\gs}{Gold Standard}
\newcommand{\silver}{silver standard}
\newcommand{\prsnr}{Pruessner}
\newcommand{\pseg}{\prsnr-protocol manual segmentation}
\newcommand{\psegs}{\prsnr-protocol manual segmentations}
\newcommand{\snt}{SNT}

% title and authors
%\title{Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using
%Multiple Automatically Generated Templates}
\title{Bootstrapping Multi-atlas Segmentation Using Multiple Automatically
Generated Templates for the Segmentation of the Whole Hippocampus and
Subfields}
\author[1]{Jon Pipitone}
\author[1]{Min Tae M. Park}
\author[1]{Julie Winterburn}
\author[1,9]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4]{Jens C. Pruessner}
\author[4,5]{Martin Lepage}
\author[1,6,9]{Aristotle N. Voineskos}
\author[1,6,7,8]{M. Mallar Chakravarty}
\author[ ]{the Alzheimer's Disease Neuroimaging Initiative\footnote{Data
used in preparation of this article were obtained from the Alzheimer's Disease
Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). As such, the
investigators within the ADNI contributed to the design and implementation of
ADNI and/or provided data but did not participate in analysis or writing of this
report. A complete listing of ADNI investigators can be found at:
http://adni.loni.ucla.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health Laboratory, Hospital for Sick
Children, Toronto, ON, Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Douglas Mental Health University Institute, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, McGill University, Montreal, QC, Canada}
\affil[6]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[7]{Institute of Biomaterials and Biomedical Engineering, University of
Toronto, Toronto, ON, Canada}
\affil[8]{Rotman Research Institute, Baycrest, Toronto, ON, Canada}
\affil[9]{Institute of Medical Science, University of Toronto, Toronto, ON, Canada}

@@

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that \multiatlas approaches improve segmentation accuracy and
precision over regular atlas-based approaches.  These approaches often rely on
a large number of such manually segmented atlases (e.g.  30-80) that take
significant time and expertise to produce. We present an algorithm, \mb{} (
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar accuracy to
\multiatlas approaches. Thus, our method acts as an accurate \multiatlas
approach when using special, hard-to-define atlases that are laborious to
construct. \\
\textbf{Method:} \mb{} works by propagating atlas segmentations to a template library,
formed from a subset of target images, via transformations estimated by non-linear
image registration. The resultant segmentations are then propagated to each
target image and fused using a label fusion method. We conduct a 10-fold Monte
Carlo cross-validation of whole hippocampal segmentation on a pool of 60 ADNI
subjects over a range of parameter settings, and a leave-one-out
cross-validation (LOOCV) of hippocampal subfield segmentation using five
high-resolution atlases. Two final experiments assess \mb{} when applied to
first episode psychosis and Alzheimer's disease populations, and \mb{}
segmentations are compared with existing automated methods (\fsl{}, \freesurfer{},
\maper{}) and biases are explored.\\
\textbf{Results:} Using nine atlases and 19 template images, \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) of 0.84 over 10-folds of Monte Carlo % TODO: update
cross-validation on 60 subjects relative to manual segmentations,
and shows significantly lower variability in DSC than \multiatlas segmentation.
In a LOOCV, \mb{} reproduces the volumes of the cornu ammonis (CA) 1; CA4/dentate
gyrus (DG); and strata radiatum (SR), strata lacunosum (SL), and strata
moleculare (SM) hippocampal subfields with a percent error in volume that is
at or lower than that produced by image resampling. \mb{} produces hippocampal
volumes in a first episode psychosis patient population that are highly
correlated with expert manual segmentation volumes (Pearson $r = 0.877, t =
16.244, p < 0.001$). Compared to \fsl{} and \freesurfer{}, \mb{} shows much smaller
fixed volume bias (within $250 mm^3$ on average) to semi-automated (\snt)
segmentations available from ADNI, as well as a conservative, rather than
exaggerated, proportional volume bias.\\ \textbf{Conclusion:} We demonstrate
that \mb{} produces accurate hippocampal segmentations using only 5 atlases over
different hippocampal definitions, disease populations, and acquisition types,
as well as showing that accurate identification of the hippocampal subfields is
possible.
\end{abstract}

@@

More recently, many groups have used multiple atlases to improve overall
segmentation accuracy (i.e. \multiatlas segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into the definitive segmentation for the target. In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such as cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations. For example \citet{Collins2010} demonstrated that only
14 atlases, selected based on highest similarity between medial temporal lobe
neuroanatomy as evaluated by normalized mutual information \citep{Studholme1999}
from a library of 80 atlases, were required to achieve accurate segmentations of
the hippocampus. Additionally, several methods have been explored for label
fusion, including the STAPLE algorithm (Simultaneous Truth And Performance Level
Estimation; \citet{Warfield2004}) that computes a probabilistic segmentation
using an expectation-maximization framework from an set of competing
segmentations, or others where a subset of segmentations can be estimated using
metrics such as the sum of squared differences in the regions of interest to be
segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library ranging between 30
\citep{Heckemann2006} and 80 \citep{Collins2010} manually segmented atlases.
This strategy has the main drawback of being inflexible as it does not easily
accommodate varying the definition of the hippocampal anatomy (such as the
commonly used heuristic of subdividing the hippocampus into head, body, and tail
\citep{Poppenk2011,Pruessner2000}). Furthermore, none of these methods have
demonstrated sufficient flexibility to accommodate atlases that are somehow
exceptional such as those derived from serial histological data
\citep{Chakravarty2006,Yelnik2007} or high-resolution MRI data that enables
robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}. Due
to the recent availability of the latter, there has been increased interest in
the use of probabilistic methods for the identification of the hippocampal
subfields on standard T1-weighted images. Our group recently demonstrated that
through use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{MallarChakravarty2012}.
The novelty of this manuscript is the extension of our \multiatlas methodology
to the hippocampus using more than a single input atlas, while simultaneously
limiting the number of inputs used during segmentation, and demonstrating that
accurate identification of the hippocampal subfields is indeed possible using
this methodology.

Of special relevance to the present work is the LEAP algorithm (Learning
Embeddings for Atlas Propagation; \citet{Wolz2010}) because of its focus on
performing \multiatlas segmentation with a limited number of input atlases.  The
LEAP algorithm is a clever modification to the basic \multiatlas strategy in
which an atlas library is grown, beginning with a set of manually labelled
atlases, by successively incorporating unlabelled target images once they
themselves have been labelled using \multiatlas techniques. The sequence in
which target images are labelled is chosen so that the similarity between the
atlas images and the target images is minimised at each step, effectively
allowing for deformations between very dissimilar images to be broken up into
sequences of smaller deformations. Although \citet{Wolz2010} begin with an atlas
library of 30 MR images, this method could theoretically work using a much
smaller atlas library.  In their validation, LEAP was used to segment the whole
hippocampus in the ADNI-1 baseline dataset, achieving a mean Dice score of 0.85
against \gs{} segmentations. %TODO: which segmentations were used? SNT?

@@

In this paper we have describe a thorough validation of the \mb{} algorithm for
segmentation of the hippocampus and its subfields. First, we addressed the feasibility and
accuracy of whole hippocampus segmentation with a limited number of input
atlases \citep{MallarChakravarty2012} by performing a multi-fold experiment
using a subset of the Alzheimer's Disease Neuroimaging Initiative (ADNI)
dataset over a range of parameters.  We then performed a leave-one-out
validation to determine if hippocampal subfields can be accurately identified
using our \multiatlas framework.  To ensure that we have not over-fit our
parameters to the aging or neurodegenerative brain, we also applied our
segmentations to a dataset of individuals suffering from first episode
psychosis.  Finally, validated our algorithm using all of the data available in
the \adnidataset{} sample and compare our segmentations to the other
segmentations available through the ADNI informatics portal.

@@

The simplest form of \multiatlas segmentation, which we call {\em basic 
\multiatlas segmentation}, involves three steps. First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propagated to the target image space. 
Third, the labels are combined into a single label by label fusion
\citep{Heckemann2006, Heckemann2011}. The basic \multiatlas segmentation method
is described in detail in other publications
\citep{Collins2010,Heckemann2011,Aljabar2009}. When only a single atlas is used,
basic \multiatlas segmentation degenerates into model-based segmentation: labels
are propagated from the atlas to a target, and no label fusion is needed.

\mb{} ({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates) bootstraps the creation of a large template library given a limited
input atlas library, and then uses the template library in basic \multiatlas
segmentation.  Images for the template library are selected from a set of input
target images, either arbitrarily or so as to reflect the neuroanatomy or
demographics of the target set as a whole (for instance, by sampling equally
from cases and controls).  The template library images are automatically
labelled by each of the atlases via label propagation.  Basic \multiatlas
segmentation is then conducted using the template library to segment the entire
set of target images (including the target images used in the construction of
the template library).  Since each template library image has multiple labels
(one from each atlas), the final number of labels to be fused for each target
may be quite large (i.e. \# of atlas $\times$ \# of templates).

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=\textwidth]{figure/MA-MAGeTBrain-Schematic}
  \end{centering}
  \caption{A schematic illustration of basic \multiatlas segmentation and \mb{}
  segmentation \label{fig:MAGeT}}
\end{figure}

@@

The following section describes for experiments conducted to assess the
segmentation quality of the \mb{} algorithm. The first two experiments assess the
validity of \mb{} using cross-validation designs. Experiment 1 investigates the
accuracy of \mb{} whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 assesses hippocampal subfield segmentation quality in a
leave-one-out cross-validation design. The last two experiments assess the
validity of the \mb{} algorithm when applied to different diseases: first episode
schizophrenia (Experiment 3), and Alzheimer's disease (Experiment 4).
Additionally, in Experiment 4, we compare \mb{} segmentations with those of
well-known automated methods and assessed segmentation bias.

@@

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using images
from the ADNI dataset \citep{Hsu2002}, and manual whole hippocampus
segmentations following the protocol laid out in \citep{Pruessner2000}.

This form of cross-validation allows us to rigorously validate a large number of
parameter settings of \mb{} (atlas and template library sizes, registration
algorithm, and label fusion method) and select the best parameters to use in
subsequent experiments.

@@

60 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset{}} standardized dataset. 20 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}. Fully manual segmentations of the
left and right whole hippocampi in these images were provided by one author
(JCP) according to the segmentation protocol specified in \citet{Pruessner2000}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI-xval-demographics, echo=F, dependson='ADNI1-xval-load', results="asis",cache=FALSE>>=
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               jens_atlases$RID, method = "reverse", test=FALSE, overall=TRUE)
caption = 
  "\\textbf{ADNI-1 cross-validation subset demographics.} 
  CN     - Cognitively Normal. 
  LMCI   - Late-onset Mild Cognitive Impairment. 
  AD     - Alzheimer's Disease.  
  Hisp   - Hispanic. 
  CDR-SB - Clinical Dementia Rating-Sum of Boxes.  
  ADAS   - Alzheimer's Disease Assessment Scale.  
  MMSE   - Mini-Mental State Examination."
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      label="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In this experiment, our dataset consists of 60 1.5T images and corresponding
\psegs{}. In each validation round, the dataset is partitioned into
a training set consisting of images and manual segmentations used as an atlas
library, and a validation set consisting of the remaining images to be segmented
by both \mb{} and \multiatlas{}. The computed segmentations are compared to the
manual segmentations (see Evaluation below).

A total of ten validation rounds were performed on each subject in the dataset,
over each combination of parameter settings. The parameter settings explored
are: atlas library size (1-9), template library size (1-20), registration method
(\ants{} or \animal{}, described below), and label fusion method (majority vote,
cross-correlation weighted majority vote, and normalized mutual information
weighted majority vote, described below).  In each validation round, both a
\mb{} and \multiatlas segmentation is produced.  A total of $10 \times 60 \times
9 \times 20 \times 2 \times 3 = \Sexpr{10*60*9*20*2*3}$ validation rounds
were conducted and resulting segmentations analysed.

@@

The labels produced by \mb{} and \multiatlas segmentation are compared to the
manual labels using the Dice similarity coefficient, and the recorded value for
each subject at each parameter setting explored in this experiment is the
average over ten validation rounds.
